Vlasov-Maxwell-Liouville informed Operator Learning for Solar Wind Modeling

### 2024-12-17 paper/main.tex
appendix and training

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{cite}

@@

@@

@@

Note that Vlasov's equation is tightly related to statistical mechanics through Liouville's theorem:

\begin{equation}
    \frac{\mathrm{df_\alpha(\mathbf{r}, \mathbf{v}, t)}}{\mathrm{dt}} = 0
\end{equation}

If we compute use the chain rule along the phase and time space, we end up with the following expression:
\begin{equation}
    \frac{\mathrm{df_\alpha(\mathbf{r}, \mathbf{v}, t)}}{\mathrm{dt}} = \frac{\partial f_\alpha}{\partial t} + \nabla_{\mathbf{r}} f_\alpha \frac{\partial \mathbf{r}}{\partial t} + \nabla_{\mathbf{v}} f_\alpha \frac{\partial \mathbf{v}}{\partial t} = 0
\end{equation}

which indeed is:

\begin{equation}
    \frac{\mathrm{df_\alpha(\mathbf{r}, \mathbf{v}, t)}}{\mathrm{dt}} = \frac{\partial f_\alpha}{\partial t} + \mathbf{v} \cdot \nabla_{\mathbf{r}} f_\alpha + \mathbf{a} \cdot \nabla_{\mathbf{v}} f_\alpha = 0
\end{equation}

@@

A cross-training process that relies on the bonds between the Maxwell and Vlasov PDE criterions will be critical; the PDE loss APORTE to the training process will be gradually incremented through epochs, and inversely correlated with the distances from the boundary contour delimiting the training space to minimize the error margin of non-representative results in the middle of the training process.

@@

\begin{align*}
    f_\alpha\left(r, v_{max}, t) &= 0 \\
    f_\alpha\left(r, v_{min}, t) &= 0 \\
    f_\alpha\left(r, v, t_0) &= 0 \\
    f_E\left(r_0, t\right) &= E_{L1}(t) \\
    f_B\left(r_0, t\right) &= B_{L1}(t) \\
\end{align*}

@@

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{cite}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}

@@

Traditional Physics informed Machine Learning relies on the usage of some fixed input grid to learn a representation of the data utterly dependant on resolution. To solve this problem, a discretization-invariant operator should be embedded in the training process, and this is pragmatically solved with the operators/functionals (mappings between function spaces) learning.

Such a general operator can be defined as a differential or integral operator. A useful way to embed operators, and take advantage of computational shortcuts at inference time, is using some kernel integral operator $\mathcal{G}$:

\begin{align*}
    \mathcal{G}[f(x); \kappa] = \int_\Omega \kappa(x, t) f(t) d(\alpha(t))
\end{align*}

for some region $\Omega$, defined kernel $k$, differentiable real valued functions $f$ and $\alpha$ such that the Riemann definition is fulfilled.

One of such operators could be the convolution, since it is defined as follows:

\begin{align*}
    \mathcal{G}[ f(x); \kappa ] = (\kappa * f)(t) = \int_{-\infty}^\infty \kappa(\tau)f(t-\tau)\tau'
\end{align*}

The original paper presenting PINOs included 3 integral operators suitable for discretization-invariant operator learning, analogous to wavelet transforms, fourier transforms, and laplace transforms \cite{li2023physicsinformedneuraloperatorlearning}.

This can be thought as a Functional analysis constraint model architecture, that pragmatically lands as a primordial Physics informed Machine Learning architecture that enables multi-resolution inference.

@@

To derive the definition of the Fourier Neural Operator, we must attend to the definition of the convolutional operator defined, and how it can be optimized under the convolution theorem.

As it is vastly known, we can demonstrate that the following is true (Look at the appendix):

\begin{align*}
    \mathcal{G}[ f(x); \kappa ] = (\kappa * f)(t) = \int_{-\infty}^\infty \kappa(\tau)f(t-\tau)\tau' = \mathcal{F}^{-1}[\mathcal{F}[f] \cdot \mathcal{F}[\kappa]]
\end{align*}

So, we can define some frequency parameter space $\theta_\omega$, for some $R \in \theta_\omega$ such that $R = \mathcal{F}[\kappa]$, we get the following:

\begin{align*}
    \mathcal{G}[ f(x); R \in \theta_\omega] = \mathcal{F}^{-1}[\mathcal{F}[f] \cdot R]
\end{align*}

Here, we defined $\mathcal{G}$ as some operators with learnable parameters $R$ within the linear space $\theta_\omega$. This is the FNO, as it appropiately extrapolates some pre-defined operator to the frequency domain for computational ease.

@@

@@

A cross-training process that relies on the bonds between the Maxwell and Vlasov PDE criterions will be critical; the PDE loss contribution to the training process will be gradually incremented through epochs, and inversely correlated with the distances from the boundary contour delimiting the training space to minimize the error margin of non-representative results in the middle of the training process.

@@

\begin{align*}
    f_p(r_0, v, t) = n_{L1}(t) \left( \frac{m_p}{2 \pi k_B T_{L1}(t)}\right)^{\frac{3}{2}}& \exp\left({-\frac{m_p | v_{L1}(t) - u_{L1}(t) |^2}{2k_BT_{L1}(t)}}  \right) \\
    f_p\left(r, v_{min}, t) &= 0 \\
    f_p\left(r, v, t_0) &= 0 \\
    f_e(r_0, v, t) = n_{L1}(t) \left( \frac{m_e}{2 \pi k_B T_{L1, \perp}(t)}\right)&\left( \frac{m_e}{2 \pi k_B T_{L1, \parallel}(t)}\right)^{\frac{3}{2}} \exp\left(-\frac{m_ev_{L1, \perp}^2}{2k_BT_{L1, \perp}} - \frac{m_e (v_{L1, \parallel} - u_{L1, \parallel})^2}{2k_BT_{L1, \parallel}}\right) \\
    f_e\left(r, v_{min}, t) &= 0 \\
    f_e\left(r, v, t_0) &= 0 \\
    f_E\left(r_0, t\right) &= E_{L1}(t) \\
    f_E\left(r, t_0\right) &= 0 \\
    f_B\left(r_0, t\right) &= B_{L1}(t) \\
    f_B\left(r, t_0\right) &= 0 \\
    f_v\left(r_0, t\right) &= V_{L1}(t) \\
    f_v\left(r, t_0\right) &= 0 \\
\end{align*}

@@

\section{Appendix}

\begin{definition}[Riemann-Stieljes integral]
    Given some partition $P = \{x_0, x_1, \ldots, x_n\}$ in $[a, b]$, we define $t_k \in [x_{k-1}, x_k]$ such that a partial sum of the form
    \begin{align*}
        S(P, f, \alpha) = \sum_{k = 1}^n f(t_k) \Delta \alpha_k
    \end{align*}
    is termed sum of Riemann-Stieljes of $f$ with respect to $\alpha$. $f$ is defined as Riemann-integrable with respect to $\alpha$ if $\exists N \in \mathbb{N}$ such that for every $\varepsilon > 0$ there is some $P_\varepsilon \subset [a, b]$ that encapsulates other partition $P$, and for every election of points $t_k \in [x_{k-1}, x_k]$, $\mid S(P, f, \alpha) - N \mid < \varepsilon$.
\end{definition}

\begin{definition}[Fourier Series]
    The Fourier series of a function $f$, periodic under the interval $\Omega : [0, T]$, is given by the following series:
    \begin{align*}
        f(x) = \sum_{n = -\infty}^\infty c_n \exp \left( i 2\pi \frac{n}{T} x\right)
    \end{align*}

    where $c_n$ is some complex number defined within the symmetric consideration $\overline{c_k} = c_{-k}$ $ \forall k \in \mathbb{N}$ that is derived as follows:
    \begin{align*}
        f(x) &= \sum_{n = -\infty}^\infty c_n  e^{i 2\pi \frac{n}{T} x} \\
        f(x) &= c_m e^{i 2\pi \frac{m}{T} x} + \sum_{n \neq m} c_n e^{i 2\pi \frac{n}{T} x} \\
        c_m e^{i 2\pi \frac{m}{T} x} &= f(x) - \sum_{n \neq m} c_n e^{i 2\pi \frac{n}{T} x} \\
        c_m &= e^{- i 2\pi \frac{m}{T} x} f(x) - e^{-i 2\pi \frac{m}{T} x} \sum_{n \neq m} c_n e^{i 2\pi \frac{n}{T} x} \\
        \int_0^T c_m dx &= \int_0^Te^{- i 2\pi \frac{m}{T} x} f(x)dx  - \int_0^T \sum_{n \neq m} c_n e^{i 2\pi \frac{n - m}{T} x} dx\\
        c_m &= \frac{1}{T} \int_0^Te^{- i 2\pi \frac{m}{T} x} f(x)dx
    \end{align*}

From this definition we can derive an interpretation of the Fourier Series as a reflection of an infinite orthogonal base with respect of the function.
\end{definition}

\begin{definition}[Fourier Transform]
    Let $f$ be a non-periodic function. We can define the Fourier transform of $f$ as an extreme case of periodicity, interpreting it as tending to infinity:
    \begin{align*}
        \mathcal{F}[f] = \int_{-\infty}^{\infty} f(x) e^{- i 2 \pi s x} dx
    \end{align*}
\end{definition}

\begin{theorem} [Convolution Theorem]
    Given the following definition of some convolution under some domain $\Omega$ of some differentiable real valued functions $f$ and $g$:
    \begin{align*}
        h(x) = \int_{\Omega} f(t) g(x - t) dt
    \end{align*}
Given that $h \in L^1\left( \mathbb{R}^n\right)$ (the demonstration is out of scope), we can assert the following:
\begin{align*}
    \mathcal{F}[h] = \int_{\Omega} \left(\int_{\Omega} f(t) g(x - t) dt \right)e^{-i 2 \pi s x} dx
\end{align*}
Using the Fubini theorem to interchange the order of integration, we have:
\begin{align*}
    \mathcal{F}[h] = \int_{\Omega} \left(\int_{\Omega} f(t) g(x - t) e^{-i 2 \pi s x} dx\right) dt
\end{align*}
and finally, we make the following substitution $y = x - t$, with its respective jacobian being 1:
\begin{align*}
    \mathcal{F}[h] &= \int_{\Omega} \left(\int_{\Omega} f(t) g(y) e^{-i 2 \pi s (y + t)} dy\right) dt \\
    \mathcal{F}[h] &= \left(\int_{\Omega} f(t) e^{-i 2 \pi s t} dt \right) \left(\int_{\Omega} g(y) e^{-i 2 \pi s y} dy \right) \\
    \mathcal{F}[h] &= \mathcal{F}[f] \cdot \mathcal{F}[g]
\end{align*}
\end{theorem}

\begin{definition}[Boltzmann's equation]
    Given a probability density function $f(\mathbf{r}, \mathbf{v}, t)$ defining the probability of finding a particle in the phase space at the volume $d^3\mathbf{r}d^3\mathbf{v}$, Boltzmann's general statement is as follows:
    \begin{align*}
        \frac{\mathrm{df}}{\mathrm{dt}} = \left( \frac{\partial f}{\partial t} \right)_{force} + \left( \frac{\partial f}{\partial t} \right)_{coll}+ \left( \frac{\partial f}{\partial t} \right)_{diff}
    \end{align*}
    the $force$ term being the change due to external interference, $diff$ the diffusion of the particles, and $coll$ the forces due to collisions.

    Now, given a fixed probability at time $t$ of some particles to be at $r$ and velocity $v$ in the volume $d^3\mathbf{r}d^3\mathbf{v}$, let's consider a shift due to a force $\mathbf{F}$ such that at time $t + \Delta t$, the particle is at $\mathbf{r} + \Delta \mathbf{r}= \mathbf{r} + \Delta t \mathbf{v}$ with velocity $\mathbf{v} + \Delta \mathbf{v} = \mathbf{v} + \frac{\mathbf{F}}{m} \Delta t$, and given Liouville's theorem, the phase space of probabilities doesn't change over time:
    \begin{align*}
        f\left(\mathbf{r} + \mathbf{v} \Delta t, \mathbf{v} + \frac{\mathbf{F}}{m} \Delta t, t + \Delta t\right) d^3\mathbf{r} d^3 \mathbf{v} = f(\mathbf{r}, \mathbf{v}, t) d^3\mathbf{r} d^3 \mathbf{v}
    \end{align*}
    But under collisions, we have to account for the change of the probability in the phase space due to the $\left( \frac{\partial f}{\partial t} \right)_{coll}$ term:
    \begin{align*}
        dN_{coll} &= \left( \frac{\partial f}{\partial t} \right)_{coll} \Delta t d^3\mathbf{r} d^3 \mathbf{v} \\
        dN_{coll} &=  f\left(\mathbf{r} + \mathbf{v} \Delta t, \mathbf{v} + \frac{\mathbf{F}}{m} \Delta t, t + \Delta t\right) d^3\mathbf{r} d^3 \mathbf{v} - f(\mathbf{r}, \mathbf{v}, t) d^3\mathbf{r} d^3 \mathbf{v} = \Delta f d^3\mathbf{r} d^3\mathbf{v} \\
        \frac{dN_{coll}}{d^3\mathbf{r}d^3\mathbf{v}} &= \frac{\Delta f}{\Delta t} \to^{\Delta f \to 0}_{\Delta t \to 0} \to \left(\frac{\partial f}{\partial t} \right)_{coll} = \frac{df}{dt}
    \end{align*}

Finally, if we expand upon the total differential of $f$ we get the following:
\begin{align*}
    df &= \frac{\partial f}{\partial t} dt + \frac{\partial f}{\partial x} dx + \frac{\partial f}{\partial y} dy + \frac{\partial f}{\partial z} dz + \frac{\partial f}{\partial v_x} dv_x + \frac{\partial f}{\partial v_y} dv_y + \frac{\partial f}{\partial v_z} dv_z \\
    df &= \frac{\partial f}{\partial t} dt + \nabla f \cdot d\mathbf{r} + \nabla_{\mathbf{v}} f \cdot d\mathbf{v} \\
    \frac{df}{dt} &= \frac{\partial f}{\partial t} + \nabla f \cdot \frac{d\mathbf{r}}{dt} + \nabla_{\mathbf{v}} f \cdot \frac{d\mathbf{v}}{dt} = \frac{\partial f}{\partial t} + \nabla f \cdot \mathbf{v}+ \nabla_{\mathbf{v}} f \cdot \frac{\mathbf{F}}{m}
\end{align*}

Getting our two conclusions with collisionless interaction of particles (Vlasov Equation) and collisions (Boltzmann Equation):
\begin{align*}
    \frac{\partial f}{\partial t} + \nabla f \cdot \mathbf{v}+ \nabla_{\mathbf{v}} f \cdot \frac{\mathbf{F}}{m} &= 0 \\
    \frac{\partial f}{\partial t} + \nabla f \cdot \mathbf{v}+ \nabla_{\mathbf{v}} f \cdot \frac{\mathbf{F}}{m} &= \left( \frac{\partial f}{\partial t} \right)_{coll} \\
.\end{align*}

\end{definition}

\begin{definition}[Lagrange multipliers]
    Given a real valued function $f: \mathbb{R}^n \to \mathbb{R}$, let's say we want to find the local extrema of this function tied to some constraints $g_i: \mathbb{R}^n \to \mathbb{R}$. We define Lagrange Multipliers $\lambda_i$, real valued scalars, as follows:
    \begin{align*}
        \nabla f = \sum_{ i } \lambda_i \nabla g_i
    \end{align*}
    this definition comes from the colinearity of the gradient of constraints with respect to the function $f$. We can also create a function named \mathbf{Lagrangian function} to make the surrogate:
    \begin{align*}
        \mathcal{L} = f + \mathbf{\lambda} \cdot \mathbf{g}
    \end{align*}
    such that the following describes the extrema under defined constraints:
    \begin{align*}
        \nabla \mathcal{L} = 0
    \end{align*}
\end{definition}
